 

### 2022-05-31 thesis/abstract.qmd
Finish thesis work

Die maximale Sauerstoffaufnahme (V̇O~2max~) ist die wichtigste physiologische Kenngröße der Ausdauerleistungsfähigkeit. Zur Bestimmung der V̇O~2max~ werden während eines Rampentests die Atemgase kontinuierlich gemessen. Die üblicherweise im breath-by-breath-Verfahren erhobenen Werte erfordern eine Weiterverarbeitung um die V̇O~2max~ zu bestimmen. Unterschiedliche Verfahren der Datenverarbeitung können die gemessene V̇O~2max~ beeinflussen. Ziel der vorliegenden Arbeit ist es, anhand eines scoping review zu bestimmen, welche Datenverarbeitungsstrategien in der wissenschaftlichen Literatur genutzt werden, um diese im Anschluss anhand experimenteller Daten zu vergleichen. Für den Review wurde eine Zufallsstichprobe von 500 Artikeln untersucht; der experimentelle Vergleich erfolgte anhand von 72 standardisierten Rampentests auf dem Laufband. Die im Review untersuchten Studien benutzen eine Vielzahl von unterschiedlichen Datenverabeitungsstrategien, welche häufig nur unzureichend beschrieben wurden. Die verschiedenen Strategien können die V̇O~2max~ im Median zu mehr als 5% beeinflussen, bei teils deutlich größeren individuellen Abweichungen. Die Ergebnisse verdeutlichen den Bedarf einer Standardisierung und korrekten Darstellung der benutzten Datenverarbeitungsstrategien zur Bestimmung der V̇O~2max~. Diese Arbeit bietet hierfür Empfehlungen zur Verwendung und Vergleichbarkeit verschiedener Strategien, sowie Richtlinien zur Beschreibung dieser im Rahmen wissenschaftlicher Studien.

>>>>

Die maximale Sauerstoffaufnahme (V̇O~2max~) ist die wichtigste physiologische Kenngröße der Ausdauerleistungsfähigkeit. Zur Bestimmung der V̇O~2max~ werden während eines Rampentests die Atemgase kontinuierlich gemessen. Die üblicherweise im breath-by-breath-Verfahren erhobenen Werte erfordern eine Weiterverarbeitung um die V̇O~2max~ zu bestimmen. Unterschiedliche Verfahren der Datenverarbeitung können die gemessene V̇O~2max~ beeinflussen. Ziel der vorliegenden Arbeit ist es, anhand eines scoping review zu bestimmen, welche Datenverarbeitungsstrategien in der wissenschaftlichen Literatur genutzt werden, um diese im Anschluss anhand experimenteller Daten zu vergleichen. Für den Review wurde eine Zufallsstichprobe von 500 Artikeln untersucht; der experimentelle Vergleich erfolgte anhand von 72 standardisierten Rampentests auf dem Laufband. Die im Review untersuchten Studien benutzen eine Vielzahl von unterschiedlichen Datenverabeitungsstrategien, welche häufig nur unzureichend beschrieben werden. Die verschiedenen Strategien können die V̇O~2max~ im Median zu mehr als 5% beeinflussen, bei teils deutlich größeren individuellen Abweichungen. Die Ergebnisse verdeutlichen den Bedarf einer Standardisierung und korrekten Darstellung der benutzten Datenverarbeitungsstrategien zur Bestimmung der V̇O~2max~. Diese Arbeit bietet hierfür Empfehlungen zur Verwendung und Vergleichbarkeit verschiedener Strategien, sowie Richtlinien zur Beschreibung dieser im Rahmen wissenschaftlicher Studien.

### 2022-05-31 thesis/discussion.qmd
Finish thesis work

This thesis aimed to map current practices of data processing strategies to determine V̇O~2max~ and to compare them on experimental data. The results show, that current research uses a variety of strategies to determine V̇O~2max~, which directly influences the values obtained. Many articles only incompletely report on their methods and use outdated strategies. These practices hinder the validity and reproducibility of V̇O~2max~ measurement.

## Current state of data processing

Despite calls to use moving averages [@robergs2003; @robergs2010], binned time averages remain the most commonly used data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies). These numbers are generally in line with the findings of the non-systematic search by Midgley et al. @midgley2007 and the survey by Robergs et al. @robergs2010. It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations, that discourage researchers from using binned averages [@robergs2010]. Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison). The peak in oxygen uptake may be attained between two averaging intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example). Binned time average revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data. Despite these arguments speaking against the use of binned time averages, my review demonstrates that they remain extremely common in the scientific literature.

@@

The length of the calculation interval for averaging is highly diverse within the literature (see @fig-duration). This may reflect contradictory recommendations [@howley1995; @robergs2003]. As long as there is no consensus on the arguments speaking for shorter or longer calculation intervals, there appears to be no optimal interval duration. As different interval durations can heavily influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.

The exact reasons for exercise scientist to ignore most recommendations by using binned time averages remain unknown. Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~. But past publications on this issue have been frequently cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field. Researchers may also use binned averages for traditional reasons. Douglas bags, as well as many mixing chamber devices, measure the oxygen uptake over fixed time intervals, producing data apparently similar to those by a binned time average of breath-by-breath data. But comparability with older data should only be an issue when using data acquired by different measurement methods within one analysis. Current studies using new breath-by-breath data do not reasonably need to rely on outdated methods of data processing.

A major source for choosing suboptimal processing strategies may be limitations by analysis software. My review shows that most researchers use the software of the metabolic cart's manufacturer to analyse the gas exchange data. These software may by default output binned time averages instead of raw breath-by-breath data. Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software. This may also explain why digital filtering has --- despite recommended by Robergs et al. @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations. The common malpractices of data processing are likely to be both attributed to a lack of awareness and a lack of adequate software solutions.

## Impact of different data processing

The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment. In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] --- and pure logic --- longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison). The analysed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019]. Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison. The evaluation of unprocessed raw data for its maximum is highly erroneous (see the individual breath data points in @fig-example) and as such is not performed in research (see @fig-duration); so there is no reason to include it in a comparison to other strategies. While previous research was often conducted in sedentary or recreational individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.

@@

Moving time and moving breath averages lead to almost identical V̇O~2max~ values on a median level (@fig-comparison). This seems natural in that the athletes in this study reached respiratory rates around 60 min^-1^ (see @fig-rr), resulting in equivalent time- and breath-based interval lengths. For an athletic population, V̇O~2max~ values obtained by moving time and moving breath average can approximately used interchangeably. Given that less trained individual display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will not translate to a sedentary population.

The exact impact of data processing strategies on the V̇O~2max~ is highly individual. Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019]. On an individual level, data processing strategies may impact the V̇O~2max~ in different severity. For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison). Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classifications. However on an individual level these can only be applied with a high margin of error. Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others. Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported. While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, on the individual level the raw data from each test is required.

## Guidelines for data processing and reporting

This thesis focussed on the occurrence and impact of different data processing strategies, but did not investigate their validity in the first place. However, the results of the scoping review and experimental comparison allow to specify existing recommendation for data processing and set new guidelines for data reporting.

I highly disregard researchers from continuing to use binned time averages to determine V̇O~2max~. The main reason for such procedure is pure tradition, as it reduces the breath-by-breath data in an inappropriate way leading to a small, yet systematic, underestimation of V̇O~2max~. Moving time or moving breath averages are preferable to binned averages. In an athletic population they may be used equivalent, but in sedentary individuals breath-based averages will lead to lower V̇O~2max~ values than time-based averages with the same interval length parameter. While using a constant breath interval will lead to a similar degree of data smoothing regardless of the training status, using a constant time interval will lead to the same physiological time-frame for determining V̇O~2max~ (but with a different degree of data smoothing). From a data processing perspective, I prefer using breath-based moving averages, while keeping in mind, that this may underestimate true V̇O~2max~ in individuals with a low respiratory rate in the end of the exercise test.

@@

To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategy is crucial. The review presented here shows that almost half of the studies that measured V̇O~2max~ did not describe their processing strategy. Other aspects of the data processing, such as outlier filtering or rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting). @tbl-recommendation lists 7 item that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~. These items may be reported in form of a checklist, as an in-text enumeration or in sentence form. An example paragraph containing all the relevant information for the original data presented in this thesis [@quittmann_unpubl_a; @quittmann_unpubl_b] would be:

> *"We measured breath-by-breath data during the ramp tests with a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America). The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro Package for R [@nolte2022]. This strategy produces similar results as that recommended by Robergs et al. @robergs2010, but does not produce a time lag."*

@@

```{r}
#| echo: false
#| tbl-cap: "Recommendations for reporting data processing strategies to determine the maximum oxygen uptake"
#| label: tbl-recommendation

rec_df <- data.frame(
  item = c("Metabolic Cart", "Measurement Mode", "Software", "Preprocessing", "Processing Strategy", "Processing Parameters", "Rationale"),
  desc = c(
    "State the exact device model and manufacturer",
    "State the measurement mode (e.g., mixing chamber, breath-by-breath,…)",
    "State the name and version of the software used for data analysis",
    "State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis",
    "State the exact data processing strategy used to determine the VO2max (e.g., binned time averages)",
    "State the parameters used for the processing strategy (e.g., length of averaging interval)",
    "State the rationale for using the processing strategy (e.g., reference to recommendations)")
)

@@

Ambiguities in the reporting of the investigated studies may impact the analysis results. For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter durations (e.g., 4x15 second), without describing this correctly. Moreover the exact definitions for building binned averages varies within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint. Additionally some studies reviewed did not define the maximum, but a preset binned average period as their V̇O~2max~ (for example the last bin, regardless of its value). In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing. I did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguity in reporting.

>>>>

This thesis aimed to review current practices of data processing strategies to determine V̇O~2max~ and to compare them on experimental data. The results show, that current research uses a variety of processing strategies to determine V̇O~2max~, which directly influences the values obtained. Identical raw breath-by-breath data can result in different V̇O~2max~ values when processed differently. Many articles only incompletely report on their methods and use outdated strategies. These practices hinder the validity and reproducibility of V̇O~2max~ measurement.

## Current State of Data Processing

Despite calls to use moving averages [@robergs2003; @robergs2010], binned time averages remain the most commonly used data processing strategy to determine V̇O~2max~ in the reviewed literature (see @fig-strategies). The found numbers are generally in line with the findings of the non-systematic search by Midgley et al. @midgley2007 and the survey by Robergs et al. @robergs2010. It is somewhat surprising, that practices have not changed in recent years despite the publication of recommendations, that discourage researchers from using binned averages [@robergs2010]. Using binned time averages leads to systematically lower V̇O~2max~ values as compared to moving averages (see @fig-comparison). The peak in oxygen uptake may be attained between two averaging intervals, resulting in an underestimation of V̇O~2max~ (see @fig-example for an example). Binned time averages revoke the most important argument for measuring breath-by-breath: the high temporal resolution of data. Despite these arguments speaking against the use of binned time averages, my review demonstrates that they remain extremely common in the scientific literature.

@@

The length of the calculation interval for averaging is highly diverse within the literature (see @fig-duration). This may reflect contradictory recommendations [@howley1995; @robergs2003]. As long as there is no consensus on the arguments speaking for shorter or longer calculation intervals, there appears to be no optimal interval duration. As different interval durations influence V̇O~2max~ by more than 5% on a median level (see @fig-comparison), the exact reporting of the data processing strategy remains essential for interpretation.

The exact reasons for exercise scientists to ignore most recommendations by using binned time averages remain unknown. Some researchers may simply not be aware of the impact different data processing strategies have on the V̇O~2max~. But past publications on this issue have been frequently cited [@robergs2010; @midgley2007; @astorino2009] and should be known to most scientists in this field. Researchers may also use binned averages for traditional reasons. Douglas bags, as well as many mixing chamber devices, measure the oxygen uptake over fixed time intervals, producing data apparently similar to those by a binned time average of breath-by-breath data. But comparability with older data should only be an issue when using data acquired by different measurement methods within one analysis. Current studies using newly aquired breath-by-breath data do not reasonably need to rely on outdated methods of data processing.

A major source for choosing suboptimal processing strategies may be limitations by analysis software. My review shows that most researchers use the software of the metabolic cart's manufacturer to analyze the gas exchange data. These software may by default output binned time averages instead of raw breath-by-breath data. Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software. This may also explain why digital filtering has --- despite recommended by Robergs et al. @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations. The common malpractices of data processing are likely to be both attributed to a lack of awareness and a lack of adequate software solutions.

## Impact of Different Data Processing Strategies

The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment. In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] --- and pure logic --- longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison). The analyzed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019]. Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison. The evaluation of unprocessed raw data for its maximum is highly erroneous (see the individual breath data points in @fig-example) and as such is not performed in research (see @fig-duration); so there is no reason to include it in a comparison to other strategies. While previous research was often conducted in sedentary or recreationally trained individuals, this thesis now presents evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes.

@@

Moving time and moving breath averages lead to almost identical V̇O~2max~ values on a median level (@fig-comparison). This seems natural in that the athletes in this study reached respiratory rates around 60 min^-1^ (see @fig-rr), resulting in equivalent time- and breath-based interval lengths. For an athletic population, V̇O~2max~ values obtained by moving time and moving breath averages can approximately be used interchangeably. Given that less trained individuals display lower respiratory rates during exercise tests to exhaustion [@blackie1991], this finding will not translate to a sedentary population.

The exact impact of data processing strategies on the V̇O~2max~ is highly individual. Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019]. On an individual level, data processing strategies may impact the V̇O~2max~ in different severity. For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison). Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparability of group results for meta-analyses or group classifications. However, on an individual level these equations can only be applied with a high margin of error. Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others. Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported. While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, on the individual level the raw data from each test is required.

## Guidelines for Data Processing and Reporting

This thesis focussed on the occurrence and impact of different data processing strategies, but did not investigate their validity in the first place. However, the results of the scoping review and experimental comparison allow to specify existing recommendation for data processing and set new guidelines for reporting.

I highly disregard researchers from continuing to use binned time averages to determine V̇O~2max~. The main reason for such procedure is pure tradition, as it reduces the breath-by-breath data in an inappropriate way leading to a small, yet systematic, underestimation of V̇O~2max~. Moving time or moving breath averages are preferable to binned averages. In an athletic population they may be used equivalent, but in sedentary individuals breath-based averages will lead to lower V̇O~2max~ values than time-based averages with the same interval length. While using a constant breath interval will lead to a similar degree of data smoothing regardless of the training status, using a constant time interval will lead to the same physiological time-frame for determining V̇O~2max~ (but with a different degree of data smoothing). From a data processing perspective, I prefer using breath-based moving averages, while keeping in mind, that this may underestimate true V̇O~2max~ in individuals with a low respiratory rate in the end of the exercise test.

@@

To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategies is crucial. The review presented here shows that almost half of the studies that measured V̇O~2max~ did not describe their processing strategy. Other aspects of the data processing, such as outlier filtering or rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting). @tbl-recommendation lists seven items that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~. These items may be reported in form of a checklist, as an in-text enumeration or in sentence form. An example paragraph containing all the relevant information for the original data presented in this thesis [@quittmann_unpubl_a; @quittmann_unpubl_b] would be:

> *"We measured breath-by-breath data during the ramp tests with a ZAN 600 USB device (nSpire Health, Inc., Longmont, CO, United States of America). The raw data was analyzed without any previous filtering by using a low-pass forward-backward Butterworth filter (each filter: 3rd order, 0.04 Hz cut-off) implemented in the spiro package for R [@nolte2022]. This strategy produces similar results as that recommended by Robergs et al. @robergs2010, but does not produce a time lag."*

@@

```{r}
#| echo: false
#| tbl-cap: "Recommendations for reporting data processing strategies to determine the maximum oxygen uptake."
#| label: tbl-recommendation

rec_df <- data.frame(
  item = c("Metabolic Cart", "Measurement Mode", "Software", "Preprocessing", "Processing Strategy", "Processing Parameters", "Rationale"),
  desc = c(
    "State the exact device model and manufacturer.",
    "State the measurement mode (e.g., mixing chamber, breath-by-breath,…).",
    "State the name and version of the software used for data analysis.",
    "State if and how data underwent any initial modification (e.g., filtering of outliers, interpolation) before analysis.",
    "State the exact data processing strategy used to determine the VO2max (e.g., binned time averages).",
    "State the parameters used for the processing strategy (e.g., length of averaging interval).",
    "State the rationale for using the processing strategy (e.g., reference to recommendations).")
)

@@

Ambiguities in the reporting of the investigated studies may impact the analysis results. For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter durations (e.g., 4x15 second), without describing this correctly. Moreover the exact definitions for building binned averages vary within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint. Additionally some studies reviewed did not define the maximum, but a pre-set binned average period as their V̇O~2max~ (for example the last bin, regardless of its value). In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing. I did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguities in their reporting.

### 2022-05-31 thesis/introduction.qmd
Finish thesis work

Performance in endurance sports is limited by the human physiology. Athletes need to supply energy to the contracting muscles for locomotion, a process that happens mainly via the oxidative phosphorylation. A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster [@holloszy1984]. The highest activity of oxidative phosphorylation on a cellular level can be approximated by measuring the maximum oxygen uptake (V̇O~2max~) on a whole-body level.

@@

Differences in V̇O~2max~ due to data processing can cause serious implications in practice. They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analysis [@martin-rincon2019]. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009]. Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].

Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. Howley et al. @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~. Based on synthetic and experimental data, Robergs & Burnett @robergs2003 opposed this conclusion, stating that shorter calculation intervals are less erroneous. They further argued to use breath-based moving averages instead of time averages. In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003]. In a classic opinion piece, Robergs et al. @robergs2010 argued for using digital filtering (i.e. a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by Weir et al. @weir2004. However, Robergs et al. @robergs2010 acknowledged the lack of accessible software implementing such procedures.

In absence of one commonly accepted method, data processing varies among the literature. In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research articles [@myers1990; @astorino2009; @scheadler2017; @johnson1998]. Midgley et al. @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals. They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all. While providing interesting first insights into reporting and processing practices, the search by Midgley et al. @midgley2007 was not systematic and its methods were not described in detail. Robergs et al. @robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*. They provided the latter one with a total of 75 respondents, who reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning. Practices in reporting and processing may have changed more than a decade after the publication of these numbers.

@@

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. Martin-Rincon et al. @martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts. Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies.

Differences in V̇O~2max~ due to data processing strategies may have serious implication for individuals [@johnson1998], yet almost all comparisons only report mean differences between strategies. No research has yet compared a variety of systematically derived strategies among a group of well-trained individuals using a standardized measurement set-up.

This thesis will first map current practices of data processing for V̇O~2max~ determination by the means of a systematic scoping review of current scientific literature. Secondly, I will compare different data processing strategies in relation to the most commonly applied on a set of 72 standardized treadmill tests in well-trained individuals. The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals. The review allows to assess the implementation of current data processing recommendations and to find malpractices in reporting. The results build a basis for providing new recommendations for data processing and its reporting for determining the V̇O~2max.~

>>>>

Performance in endurance sports is mainly limited by the human physiology. Athletes need to supply energy to the contracting muscles for locomotion, a process that happens mainly via the oxidative phosphorylation. A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster [@holloszy1984]. The highest activity of oxidative phosphorylation on a cellular level can be approximated by measuring the maximum oxygen uptake (V̇O~2max~) on a whole-body level.

@@

Differences in V̇O~2max~ due to data processing can cause serious implications in practice. They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analyses [@martin-rincon2019]. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009]. Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].

Calls to standardize processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. Howley et al. @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~. Based on synthetic and experimental data, Robergs & Burnett @robergs2003 opposed this conclusion, stating that shorter calculation intervals are less erroneous. They further argued to use breath-based moving averages instead of time averages. In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003]. In a classic opinion piece, Robergs et al. @robergs2010 argued for using digital filtering (i.e. a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by Weir et al. @weir2004. However, Robergs et al. @robergs2010 acknowledged the lack of accessible software implementing such procedures.

In absence of one commonly accepted method, data processing varies among the literature. In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research [@myers1990; @astorino2009; @scheadler2017; @johnson1998]. Midgley et al. @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals. They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all. While providing interesting first insights into reporting and processing practices, the search by Midgley et al. @midgley2007 was not systematic and its methods were not described in detail. Robergs et al. @robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*. They provided the latter one with a total of 75 respondents, who reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning. Practices in reporting and processing may have changed more than a decade after the publication of these numbers.

@@

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. Martin-Rincon et al. @martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary individuals and recreational athletes, using two different metabolic carts. Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies.

Differences in V̇O~2max~ due to data processing strategies may have serious implication for individuals [@johnson1998], yet almost all comparisons only report mean differences between strategies. No research has yet compared a variety of systematically derived strategies among a group of trained individuals using a standardized measurement set-up.

This thesis will first map current practices of data processing for V̇O~2max~ determination by the means of a systematic scoping review of current scientific literature. Secondly, I will compare the most common data processing strategies on a set of 72 standardized treadmill tests in well-trained individuals. The results will help to compare V̇O~2max~ data derived from different processing methods among studies and in individuals. The review allows to assess the implementation of current data processing recommendations and to find malpractices in reporting. The results build a basis for providing new recommendations for data processing to determine the V̇O~2max~ and its reporting in research articles.
